\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

First and foremost, I would like to thank my two supervisors, Pierre Henneaux and Pierre-Etienne Labeau. They have been following me since my master's and played a big role in this thesis. I particularly enjoyed the autonomy they gave me, their knowledge in power systems and reliability, and I am still impressed by Pierre-Etienne always managing to provide me with relevant comments and suggestions despite meetings which (by my fault) have sometimes been a bit too infrequent.

I would also like to thank the members of my committee and of the jury for taking the time to assess this thesis, for their comments before, during, and after the defence, for widening my perspectives, and for bringing this thesis closer to something that can be used by utilities.

I had the chance to perform my thesis in the framework of a joint project between Université de Liège, ULB, KULeuven and Haulogy (then Blacklight Analytics), the CYPRESS project. This allowed me to have interactions with many researchers from different institutions which is very welcome in what can sometimes be a solitary adventure. I would particularly like to thank Louis for initiating the project, Efthymios and Shahab for their initiative and for pushing the group together, Jean-Michel, Sami, and Vincent for helping me understand how cyber systems work and for willing to learn how power systems work, and the SPF Économie for funding the project.

I would also like to thank Panos and Keith for inviting me to stay three months at Strathclyde, Ifigeneia and Sam for our common work, Luke, Susan, Shanay, and Hengqing for being at the office almost every day in a post-covid period, Brian for hosting me, for the guitar and the barbecues, and everyone else at Strathclyde for the warm welcome and the Paesano's pizzas. I don't remember all your names (I have never been good at this), but I remember your company.

I also need to thank the \Dynawo{} developers at RTE for making their software freely available and for helping me to make it work in all my niche applications.

Of course, my time at ULB would have never been this enjoyable without all my current and past colleagues of the BEAMS-Electrical Energy department. To name but those who have shared my office (there is no way I will risk trying to list everyone), Neriton who helped me to get started with teaching, Gilles who, with Christophe, helped me to tame the RED, Wilson who agreed that lunch starts at 11:30 or even before, Mohammad who was always willing to share his knowledge and some laughs, Alexandre who managed to have a longer commute than I (most of the time), Lionel who initially ``did not understand anything of what I said'' and helped me put some pedagogy in my presentations, and Jean-Luc who will soon invite us to a nice Belgian restaurant.

And last but not least, I would like to thank my parents, my friends, and even my little sister for their continuous support for all these years.